# Judging Quality of Excersice Execution from Accelerator data


Practical Machine Learning - Course Project Writup

Mike Wise - 20 Aug 2014 - predmachlearn-004

This looks at the practicality of juding he quality of excersize execution (weight lifting) from accerlator data gathered from sensors located on 
the subjects body.


# Data Processing
## Loading and preprocessing the data
Here we link in some libraries will will be using in the following code, we presume the directory
has already been set to contain the data.
We then load the data (as a data.table as opposed to a data.frame 
for faster and more flexible manipluation)
## Load the data
Here we load the data 
```{r}
library(data.table)
library(ElemStatLearn)
library(randomForest)
library(caret)

# Data Processing
## Load the data

otrn <- data.table(read.csv("pml-training.csv"))
otst <- data.table(read.csv("pml-testing.csv"))
```

There are 160 columns in the original data, many of them blank and filled with NA values. We reduce the dataset, throwing
away all columns that are not numeric or integer. We also throw away the column "num_window" as that is not sensor data.

```{r}
## Get rid of all non-numeric columns not present in both datasets
cnstst <- colnames(otrn)[7:59]
lntrn <- length(otrn)
lntst <- length(otst)

ntrn <- data.table(user_name=otrn[["user_name"]],classe=otrn[["classe"]])
ntst <- data.table(user_name=otst[["user_name"]])

for (i in 1:length(cnstst))
{
  cn <- cnstst[[i]]
  if (cn=="num_window") next
  clstst <- class(otst[[cn]])
  clstrn <- class(otrn[[cn]])
  if (clstst!="numeric" && clstst !="integer")
  {
    print(sprintf("%d tst %s is not numeric/integer but %s",i,cn,clstst))
    next
  }
  if (clstrn!="numeric" && clstrn !="integer")
  {
    print(sprintf("%d trn %s is not numeric/integer but %s",i,cn,clstrn))
    next
  }
  ntrn[[cn]] <- otrn[[cn]]
  ntst[[cn]] <- otst[[cn]]
}
```

Here we check the quality of the data, showing what columns we have retained (all relevant sensor data columns) and 
showing that we have reduced 

```{r}
# Check the quality
summary(ntrn)

ona <- sum(is.na(otrn))
nna <- sum(is.na(ntrn))
msg <- sprintf("Original training na count:%d  - After processing:%d",ona,nna)
print(msg)
```

# Model Fitting

We fit a models to the data using Random Forests. The accuracy was 100 percent.

# Random Forests
```{r}
## Random Forests
set.seed(2718)
rffit <- randomForest(classe ~ ., ntrn, importance=T)

prftrn <- predict(rffit, ntrn)
confusionMatrix(prftrn, ntrn$classe)

varImpPlot(rffit)

prftst <- predict(rffit, ntst)
prftst
```
