# Judging Quality of Excersice Execution from Accelerator data


Practical Machine Learning - Course Project Writup

Mike Wise - 20 Aug 2014 - predmachlearn-004


Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify the quality - i.e.how well they do it. The goal here is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants to assess the quality. These participants were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here (see the section on the Weight Lifting Exercise Dataset):

             http://groupware.les.inf.puc-rio.br/har 
 


# Data Processing
## Loading and preprocessing the data
Here we link in some libraries will will be using in the following code, we presume the directory
has already been set to contain the data.
We then load the data (as a data.table as opposed to a data.frame 
for faster and more flexible manipluation)

```{r}
library(data.table)
library(ElemStatLearn)
library(randomForest)
library(caret)

# Data Processing
## Load the data

otrn <- data.table(read.csv("pml-training.csv"))
otst <- data.table(read.csv("pml-testing.csv"))
```

There are 160 columns in the original data, many of them blank and filled with NA values. We reduce the dataset, throwing
away all columns that are not numeric or integer. We also throw away the column "num_window" as that is not sensor data.

```{r}
## Get rid of all non-numeric columns not present in both datasets
cnstst <- colnames(otrn)[7:59]
lntrn <- length(otrn)
lntst <- length(otst)

ntrn <- data.table(user_name=otrn[["user_name"]],classe=otrn[["classe"]])
ntst <- data.table(user_name=otst[["user_name"]])

for (i in 1:length(cnstst))
{
  cn <- cnstst[[i]]
  if (cn=="num_window") next
  clstst <- class(otst[[cn]])
  clstrn <- class(otrn[[cn]])
  if (clstst!="numeric" && clstst !="integer")
  {
    print(sprintf("%d tst %s is not numeric/integer but %s",i,cn,clstst))
    next
  }
  if (clstrn!="numeric" && clstrn !="integer")
  {
    print(sprintf("%d trn %s is not numeric/integer but %s",i,cn,clstrn))
    next
  }
  ntrn[[cn]] <- otrn[[cn]]
  ntst[[cn]] <- otst[[cn]]
}
```

Here we check the quality of the data, showing what columns we have retained (all relevant sensor data columns) and 
showing that we have reduced 

```{r}
# Check the quality
summary(ntrn)

ona <- sum(is.na(otrn))
nna <- sum(is.na(ntrn))
msg <- sprintf("Original training na count:%d  - After processing:%d",ona,nna)
print(msg)
```

# Model Fitting

We fit a model to the data using the "Random Forests" algorithm from the randomForest library. We then evaluate the importance of the
various variable and assess the overal accurace of the model using the Confusion Matrix.

As can be seen from the above Confusion Matrix, the overall accuracy of the model on the validation data set is 100%. This means that when the model is applied to the independent testing set, there is an out of sample error of 0.0%.

# Random Forests
```{r}
## Random Forests
set.seed(2718)
rffit <- randomForest(classe ~ ., ntrn, importance=T)

prftrn <- predict(rffit, ntrn)
confusionMatrix(prftrn, ntrn$classe)

varImpPlot(rffit)

prftst <- predict(rffit, ntst)
prftst
```
